---
title: "RaMWAS parameters"
author: "Andrey A Shabalin"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true # table of content true
vignette: >
  %\VignetteIndexEntry{RaMWAS parameters}
  %\VignetteEngine{knitr::rmarkdown}
---

# Initialize RaMWAS Parameters

There are several ways to initialize the parameters for calling any RaMWAS pipeline function.
The parameters can be stored in an R list like this:

```{r eval=FALSE}
param = list(
	dirbam = "/ramwas_project/bams/",
	dirproject = "/ramwas_project/",
	filebamlist = "/ramwas_project/000_list_of_files.txt",
	scoretag = "AS",
	minscore = 100,
	cputhreads = 8,
	filecpgset = "/ramwas/cpgsets/cpgset_hg19_SNPS_at_MAF_0.05.rds",
	maxrepeats = 3,
	maxfragmentsize=200,
	minfragmentsize=50,
	bamnames = NULL,
	filebam2sample = "/ramwas_project/bam2sample.txt",
	minavgcpgcoverage = 0.3,
	minnonzerosamples = 0.3,
	tbuffersize = 10e9,
	covfile = "cov.txt",
	chrkeep = 1:22
);
```

Alternatively, the parameters can be set in a separate R code file, which is processed into a list as above by `parametersFromFile` function.
The R code file can contain lines like this:

```{r eval=FALSE}
### R parameter file
dirbam = "/ramwas_project/bams/"
dirproject = "/ramwas_project/"
filebamlist = "/ramwas_project/000_list_of_files.txt"
scoretag = "AS"
minscore = 100

### platform dependent part
if(.Platform$OS.type == "windows") {
	filecpgset='C:/RaMWAS/CpG_set/cpgset_hg19_SNPS_at_MAF_0.05.rds'
} else {
	filecpgset='/computing_cluster/ramwas/cpgset_hg19_SNPS_at_MAF_0.05.rds'
}


```

# Explanation of all parameters



## Step 1, scan BAM files, record read start sites and QC information.

**Overview.** The first step of the RaMWAS pipeline includes scanning BAM files, filtering reads and recording their start locations, and calculating the QC measures.
See the separate vignette on the QC measures.

**R function.** Step 1 can be run with a single call of the `ramwas1scanBams` or separate runs of `pipelineProcessBam` for each BAM file.
The `ramwas1scanBams` function starts a local cluster with the number of jobs equal to the number of CPU cores, unless it is set by the `cputhreads` parameter.
The `pipelineProcessBam` can be used if the user wants to manually spread the task across the nodes of a computing cluster.

**Read filtering.** The alignments in the BAM files are usually filtered by the MAPQ or AS tag (set by `scoretag`), the reads with the score tag below `minscore` are discarded.

**BAM files.** The list of BAMs can specified either in the character vector `bamnames` or listed in the `filebamlist` file, one bam name per file.
The BAM file names can include full path, or path relative to `dirbam` or no path (BAMs should be in `dirbam` then).

**Files produced.** The information extracted from the BAMs is stored in R format in 'rds_rbam' subdirectory of the Project directory. The QC information in R format is stored in 'rds_qc' subdirectory of the Project directory. The QC plots and figures are stored in 'qc' subdirectory of the Project directory.














